Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models

### 2020-06-10 analysis/paper/paper.Rmd
edited the habitat covariate selection section to refer to the appendix. Ensured that all methods that used the natural log of fish density said so. Added a paragraph to the extrapolation model discussion

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat). Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points).

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with observed parr densities. We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011]. The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.

Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring volume or area were scaled to the wetted area of each site. Within each category, metrics were ranked according to their MIC value (**Figure ??**). Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.). Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.

***Worth putting the whole MIC habitat selection approach in an appendix or supplemental material, so we can include large figures and tables?***

@@

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012]. We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.

@@

We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m). We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average august temperature, wetted width:depth ratio, percent fines less than 6mm, total fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).

@@

<!-- Do we need any more in this section? I commented out the paragraph below... -->

>>>>

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. The CHaMP program generated more than 100 habitat metrics at each site, many of which were correlated with each other to one degree or another, as is often the case with stream habitat variables. We sought to include a small set of covariates that were not overly redundant (i.e. not highly correlated), described many aspects of stream habitat (e.g. substrate, temperature, complexity, etc.) and were highly associated with the observed fish densities, presumeably because they contained information about what types of habitat fish sought or avoided. Full details of how covariates were selected can be found in Appendix S1.

@@

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for spring/summer Chinook salmon parr, during summer months, using the natural log of fish densities as the response. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012]. We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.

@@

We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m). We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average august temperature, wetted width:depth ratio, percent fines less than 6mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).

@@

One of the potential downsides to the extrapolation approach used here is that the globally available attributes generally do not change through time, and therefore may not reflect the dynamic nature of changing stream habitat. While the QRF model itself uses habitat characteristics that can be observed to change over the course of several years, most GAAs are static, generally derived remotely or from another model. This is the nature of extrapolating to such large spatial extents; it can be impossible to gather actual habitat data on such a scale. 
<!-- Do we need any more in this section? I commented out the paragraph below... -->
